Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-10-23 Manuscript/manuscriptPGPE.tex
Update text and figures and add POPG data from Tiago Ferreira send by email

\section{Methods}
\subsection{Experimental C--H bond order parameters}% from the natural abundance $^{13}$C\,NMR}

The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16,antila19}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
\todo{The rest of the details to be written. I am not sure how much we need to repeat the NMRlipidsIVps paper.}

@@

Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}, with
the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
as the communication platforms.
The simulated systems are listed in 
Tables \ref{systems} (pure PE and PG bilayers without additional ions).
and \ref{systemsII} (mixtures and systems with additional ions).
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

  }\label{systems}

@@

@@

      \begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PG lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsII}
  \begin{minipage}[t]{\textwidth}
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & NaCl (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPG/K$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}    &0         & 118& 4110   &0    & 298  & 100 & 100 & \cite{CHARMM36popg}  \\
      POPG             & CHARMM36 \cite{??}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
      POPG             & CHARMM36 \cite{??}        & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGcharmm} \\
      \hline
      POPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 10664   &0     & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
      DPPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 11232  &0     & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
      DPPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 11232   &0     & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
      POPG         & Slipids \cite{??} \todoi{Ion parameters?}       & 0    & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGslipids} \\
      POPG         & Slipids \cite{??} \todoi{Ion parameters?}       & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
      \hline
      POPG             & LIPID17 \cite{??}         & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGlipid17} \\
      POPG             & LIPID17 \cite{??}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
      \hline
      POPG             & GROMOS-CKP \cite{??}         & 0              & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGgromosCKP} \\
      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
    \end{tabular}
  \end{minipage}
  %\end{sidewaystable*} 
\end{table*}

@@

  }\label{systemsII}

@@

  \todo{Data for POPC:POPG mixtures by listed by Antonio Peon is missing from this table}

@@

\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}

The glycerol backbone and $\alpha$-carbon peaks in INEPT spectra of POPE were assigned based on
previously measured POPC spectra (Fig. \ref{POPEspectra}) \cite{ferreira13}. The $\beta$-carbon peak was assigned based
on $^{13}$C chemical shift table for amines available at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
The order parameters for the glycerol backbone and headgroup C--H bonds were determined
from 2D-RPDLF and S-DROSS experiments (Fig. \ref{POPEspectra}), as described previously \cite{antila19}.
The POPE experiments were recorded at 310~K, where the bilayer is in liquid disordered phase \cite{??}.
\todo{Details to be checked by Tiago}.

\todo{Figure and discussion about POPG experiments to be addded.}

The headgroup and glycerol backbone order parameters of PE lipids are similar with 
different acyl chains and also close the values for POPC, althought PE gives systematically
slightly more positive values (Fig. \ref{HGorderParameters}). These could be explained with slightly larger temperature
in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
more positive value is farther away from zero. For PG lipids, the glycerol backbone order
parameters are more positive than for other lipids. The headgroup $\alpha$-carbon gives
value close to PE, while the value of $\beta$-carbon is distinct from other lipid being
only one which has positive sign, suggesting distinct conformation of PG lipids in this region.
This was not observed in previous $^2$H NMR study, where sign was not measured and $\beta$-carbon
order parameter was apparently similar to the value for PE and PC results.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
  \caption{\label{HGorderParameters}
    (top) Chemical structure of different lipids
    (bottom) Headgroup and glycerol backbone order parameters measured from lipids
    with different headgroups in lamellar liquid disordered phase.
    The values and signs for POPE (310~K), POPG (298~K). POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    are measured using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
  }
\end{figure}

In conclusion, the results suggests that the glycerol backbone conformations in all lipids are
relatively similar. Also, the headgroup conformations are similar for PC and PE lipids, while
PS and PG are singnificantly different. For PS lipids, the differences are discussed previously \cite{??}.

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
As reported previously for PC \cite{botan15} and PS \cite{antila19} lipids,
the headgroup and glycerol backbone order parameters show wide variation between different force fields
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
and none of the force fields reproduce all values within experimental error bars.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%\todo{Berger results are not here yet, but we should mention the ring like structures
%  pointed out by T. Piggot in the blog:
The poor performance of headgroup order parameters in Berger model can be probably explained by ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04},
which is a typical feature for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}
\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure*}
Also in line with previous studies for PC \cite{botan15} and PS \cite{antila19} lipids,
CHARMM36 simulations are in best agreement with experiments for headgroup and glycerol backbone
order parameters, and seem to capture all the essential differences between different headroups
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
%headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as previously observed also for PS lipids \cite{antila19}.
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon is PS headgroup than in PC \cite{antila19}.
In this work, the CHARMM36 simulations reproduce also the other essential differences in experimental
headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
in individual segments:
The PE headgroup order parameters in CHARMM36 simulations are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, in contrast to the negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}). While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
significantly different in PS and PC headgroups in previous work \cite{antila19}, essential differences
between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

\subsection{PC headgroup interactions with PE and PG}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Data for CHARMM without ions to be added once this issue is solved: https://github.com/NMRLipids/MATCH/issues/83}
  %\todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}
According to the electrometer concept, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}) \cite{seelig87, scherer87,antila18}. 
The different responses of PC headgroup order parameters to PE and PG are roughly
reproduced in CHARMM36 simulations, although the changes are slightly
overestimated upon addition of PE and underestimated upon addition of PG (Fig.~\ref{HGorderparametersPCvsPEPG}).
Interestingly, the response of POPC headgroup order parameters to POPE is
close to experiments also in Berger-OPLS simulations, even though the response
of Berger force field to cholesterol was significantly overestimated in our previous work \cite{botan15}.
In all force fields except Slipids, the $\alpha$-carbon order parameters of different
hydrogens are responsing differently when mixed with PE or PG lipids \todo{Maybe we should figure out what is the reason for this}.
\todo{Maybe we should analyze the P-N vector angle from different simulations}.

>>>>

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
POPE and POPG powder were purchased from Avanti polar lipids.
The NMR experiments were identical to our previous work~\cite{antila19}.
\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}

@@

Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}, with
the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
as the communication platforms.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
\todo{Maybe we should also shortly discuss here about the reasons for slight dependence of order parameter values on the method used to reconstruct hydrogens?}\\
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

  }\label{systemsPE}

@@

  \todo{We need citations for the force fields.}

@@

      \begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PG lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsPG}
  \begin{minipage}[t]{\textwidth}
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & NaCl (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPG/K$^+$  & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPG}    &0         & 118& 4110   &0    & 298  & 100 & 100 & \cite{CHARMM36popg}  \\
      POPG             & CHARMM36 \cite{??}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
      POPG             & CHARMM36 \cite{??}        & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGcharmm} \\
      \hline
      POPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 10664   &0     & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
      DPPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 11232  &0     & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
      DPPG/Na$^+$  & Slipids \cite{jambeck13}    &0         & 288 	& 11232   &0     & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
      POPG         & Slipids \cite{??} \todoi{Ion parameters?}       & 0    & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGslipids} \\
      POPG         & Slipids \cite{??} \todoi{Ion parameters?}       & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
      \hline
      POPG             & LIPID17 \cite{??}         & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGlipid17} \\
      POPG             & LIPID17 \cite{??}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
      \hline
      POPG             & GROMOS-CKP \cite{??}         & 0              & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGgromosCKP} \\
      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
    \end{tabular}
  \end{minipage}
  % \end{sidewaystable*}
    \todo{We need citations for the force fields.}
\end{table*}

@@

  }\label{systemsMIX}

@@

  \todo{Data for POPC:POPG mixtures by listed by Antonio Peon is missing from this table} \\
    \todo{We need citations for the force fields.}

@@

\subsection{Headgroup and glycerol backbone order parameters of POPE and POPG from $^{13}$C NMR}
Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
using the 
%order parameters The order parameters for the glycerol backbone and headgroup s were determined
2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.
For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13} and
the $\beta$-carbon peak was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm} (Fig. \ref{POPEspectra}).
For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13}, while $\alpha$ and  $\gamma$-carbon peaks
\todo{How were these assigned?} (Fig. \ref{POPGspectra}). The numerical value of the $\beta$-carbon
order parameter could not be determined, because its peak overlapped with the g$_2$ peak from glycerol backbone in POPG.
However, the order parameter of $\beta$-carbon is expected to be clearly smaller than for g$_2$
based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85}.
Therefore, the beginning of the S-DROSS curve gives the sign for g$_2$ order parameter and end for $\beta$ (Fig. \ref{POPGspectra} (E)).
This is confirmed with SIMPSON calculations using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
\todo{Details to be checked by Tiago}.

As discussed previously for PC and PS headgroups \cite{ollila16,antila19}, also 
the headgroup and glycerol backbone order parameters of PE and PG are essentially independent of 
acyl chain composition, and therefore manifest mainly headgroup chemistry (Fig. \ref{HGorderParameters}).
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. 
Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
The $\alpha$-carbon order parameter of PG is similar to PE and PG, while the positive value of $\beta$-carbon
is distinct from the other lipids. Notably, this difference was not observed in previous $^2$H NMR experiments,
because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and the order parameter
signs were not measured~\cite{wohlgemuth80,gally81,borle85}.

In conclusion, the order parameter experiments suggest that the glycerol backbone conformations in all lipids
and the headgroup conformations in PC and PE lipids are relatively similar, while 
PS and PG headgropus exhibit distinct conformational sampling.
The details of sampled conformation are difficult to deduce from order parameters only,
but the distinct headgroup order parameters
of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
  \caption{\label{HGorderParameters}
    (top) Chemical structure of different lipids.
    (bottom) Headgroup and glycerol backbone order parameters
    from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
  }
  \todo{The bottom figure could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure}

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
Headgroup and glycerol backbone order parameters of PE and PG lipids
show wide variation between different force fields
and none of the force fields reproduce all values within experimental error bars.
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
The poor performance of headgroup order parameters in Berger model can be probably explained by
ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
The poor performance of glycerol backbone of Slipids simulations is systemically observed also
for other lipids in previous studies \cite{botan15,antila19}.
\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?}

Without further discussion about poorly performing force fields, we focus on more detailed analysis
of CHARMM36 simulations, which give the best overall agreement with experiments for
the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
not within the experimental error bars, it seems to capture the essential differences between different headroups.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon in PS headgroup than in PC \cite{antila19}.
In this work,
%the CHARMM36 simulations reproduce also the other essential differences in experimental
%headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
%in individual segments:
the PE headgroup order parameters are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, while negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}).
\todo{The structures are already in Fig. \ref{structures}, but we need also the dihedral distributions to finish the discussion}.
%While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
%significantly different in PS and PC headgroups in previous work \cite{antila19}, essential differences
%between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
%\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

\todo{Maybe we should analyze also some other structural features, at least P-N angles.}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}
\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure*}

\subsection{PC headgroup interactions with PE and PG}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Data for CHARMM without ions to be added once this issue is solved: https://github.com/NMRLipids/MATCH/issues/83}
  %\todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The different responses of PC headgroup order parameters to PE and PG are roughly
reproduced in CHARMM36 simulations, although the changes are slightly
overestimated upon addition of PE and underestimated upon addition of PG (Fig.~\ref{HGorderparametersPCvsPEPG}).
Interestingly, the response of POPC headgroup order parameters to POPE is
close to experiments also in Berger-OPLS simulations, even though the response
of Berger force field to cholesterol was significantly overestimated in our previous work \cite{botan15}.
In all force fields except Slipids, the $\alpha$-carbon order parameters of different
hydrogens are responsing differently when mixed with PE or PG lipids \todo{Maybe we should figure out what is the reason for this}.
\todo{Maybe we should analyze the P-N vector angle from different simulations}.

### 2019-10-23 Manuscript/manuscriptPGPEsuppl.tex
Update text and figures and add POPG data from Tiago Ferreira send by email

\begin{figure}[]
%  \centering
  \includegraphics[width=\textwidth]{../Figs/POPEexperiment.pdf}
  \caption{\label{POPEspectra}
    (A) Chemical structure of POPE with the labeling of headgroup and glycerol backbone carbons.
    (B) INEPT spectra from POPE sample with the headgroup and glycerol backbone peaks labeled.
    (C) 2D R-PDLF spectra
    (D) Dipolar sliced from the  2D R-PDLF spectra with the resulting order parameters on top of figures.
    (E) Experimetal S-DROSS curves giving signs of the order parameters.
  }
  \todo{A, B etc. labels to be put in the figure.}
\end{figure}

>>>>

\begin{figure}[]
%  \centering
  \includegraphics[width=0.8\textwidth]{../Figs/POPEexperiment.pdf}
  \caption{\label{POPEspectra}
    (A) Chemical structure of POPE with the labeling of headgroup and glycerol backbone carbons.
    (B) INEPT spectra from POPE sample with the headgroup and glycerol backbone peaks labeled.
    (C) 2D R-PDLF spectra
    (D) Dipolar sliced from the  2D R-PDLF spectra with the resulting order parameters on top of figures.
    (E) Experimetal S-DROSS curves giving signs of the order parameters.
  }
  \todo{A, B etc. labels to be put in the figure.}
\end{figure}

\begin{figure}[]
%  \centering
  \includegraphics[width=0.8\textwidth]{../Figs/PGexpRPDLF.eps}
  \caption{\label{POPGspectra}
    (A) Chemical structure of POPG with the labeling of headgroup and glycerol backbone carbons.
    (B) INEPT spectra from POPG sample with the headgroup and glycerol backbone peaks labeled.
    (C) 2D R-PDLF spectra
    (D) Dipolar sliced from the  2D R-PDLF spectra with the resulting order parameters on top of figures.
    (E) Experimetal S-DROSS curves giving signs of the order parameters.
  }
\end{figure}

\begin{figure}[]
%  \centering
  \includegraphics[width=0.8\textwidth]{../Figs/SIMPSON.pdf}
  \caption{\label{POPGsimpson}
    Simpson simulaton of S-DROSS curve of $\beta$-carbon of POPG.
  }
\end{figure}
